---
title: "Portfolio Optimization"
author: "Chloe Bouton"
date: "2025-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load Required Packages

```{r}
library(quantmod)                    
library(PerformanceAnalytics) 
library(PortfolioAnalytics)   
library(ROI)                 
library(ROI.plugin.quadprog)  
library(ROI.plugin.glpk)
library(ROI.plugin.symphony)
library(CVXR)
library(corrplot)  
```

- **quantmod** for financial data retrieval
- **PerformanceAnalytics** for performance measurement
- **PortfolioAnalytics** for portfolio optimization framework
- **ROI.plugin.quadprog** for quadratic programming solver

## Data Acquisition and Preprocessing

``` {r}
# Define ticker symbols
tickers <- c("AAPL", "MSFT", "GOOGL", "JPM", "XOM")

# Get 5 years of monthly price data
getSymbols(tickers, 
           src = "yahoo", 
           from = Sys.Date() - 365*5)

prices <- do.call(merge, lapply(tickers, function(sym) Ad(get(sym))))
colnames(prices) <- tickers

returns <- na.omit(Return.calculate(prices))
```
- **Data:** 5 years of monthly adjusted closing prices from Yahoo Finance.

- **Returns:** Calculated as log returns, cleaned for NAs.


## Performance Analysis and Risk Metrics

```{r}
# Performance summary
charts.PerformanceSummary(returns, 
                          main = "Daily Returns",
                          legend.loc = "topleft")

# Risk metrics
table.Stats(returns)
table.AnnualizedReturns(returns)
table.DownsideRisk(returns)

VaR(returns, p = 0.95, method = "historical")
ES(returns, p = 0.95, method = "historical")
```

## Correlation Analysis

```{r}
# Correlation matrix
cor_matrix <- cor(returns)
corrplot::corrplot(cor_matrix, 
                   method = "circle", 
                   type = "upper")

# Covariance matrix
cov_matrix <- cov(returns)
```
**Microsoft**, **Google** and **Apple** are the assets with the highest correlation while **JPMorgan Chase & Co** and **Exxon Mobil Corp** have little to no correlation with other stocks. We observe that all chosen stocks have positive correlation, meaning they will tend to move together and not opposite to each others.


## Portfolio Optimization Setup

```{r}
# Define Portfolio
port_spec <- portfolio.spec(assets = colnames(returns))

# Add constraints
port_spec <- add.constraint(portfolio = port_spec, 
                            type = "full_investment")

port_spec <- add.constraint(portfolio = port_spec, 
                            type = "long_only")

# Add objectives
port_spec <- add.objective(portfolio = port_spec, 
                           type = "risk", 
                           name = "StdDev")

port_spec <- add.objective(portfolio = port_spec, 
                           type = "return", 
                           name = "mean")
```

## Optimization Results and Efficient Frontier

```{r}
# Optimization
opt_result <- optimize.portfolio(R = returns, 
                                 portfolio = port_spec, 
                                 optimize_method = "ROI", 
                                 trace = TRUE)

opt_result
chart.Weights(opt_result)

# Equal-weight portfolio returns
equal_weights <- rep(1 / ncol(returns), ncol(returns))
ew_returns <- Return.portfolio(returns, 
                               weights = equal_weights)

# Optimized portfolio returns
opt_weights <- extractWeights(opt_result)
opt_returns <- Return.portfolio(returns, 
                                weights = opt_weights)

# Combine and plot
combined <- merge.xts(ew_returns, opt_returns)
colnames(combined) <- c("Equal Weight", "Optimized")
charts.PerformanceSummary(combined, main = "Performance: Equal Weight vs Optimized")


# Efficient frontier
ef <- create.EfficientFrontier(returns, 
                               port_spec, 
                               type = "mean-StdDev")

chart.EfficientFrontier(ef, 
                        match.col = "StdDev", 
                        n.portfolios = 25, 
                        main = "Efficient Frontier")
```
Given the correlations observed earlier, it is no surprise that only one stock among Apple, Microsoft, and Google was selected for the optimized portfolio, alongside the two remaining stocks, JPM and XOM.

Reviewing the annualized returns and volatility calculated at the beginning:

AAPL:
- annualized return: 0.2459
- volatility: 0.2986
- Sharpe ratio: 0.8236

MSFT:
- annualized return:0.1831
- volatility: 0.2696
- Sharpe ratio: 0.6793

GOOGL:
- annualized return:0.1936
- volatility: 0.3087
- Sharpe ratio: 0.6270

AAPL’s higher annualized return and Sharpe ratio, compared to its counterparts, explain why it was chosen among the three highly correlated stocks.



## Risk Contribution Analysis

```{r}
opt_weights <- as.numeric(opt_weights)

# Portfolio volatility 
port_volatility <- sqrt(sum(opt_weights * (cov_matrix %*% opt_weights)))

# Risk contribution for each asset
risk_contrib <- (cov_matrix %*% opt_weights) * opt_weights / port_volatility

# Convert to a data frame
risk_contrib_df <- data.frame(Asset = colnames(returns), RiskContribution = risk_contrib)

print(risk_contrib_df)
```
The risk contribution table can be directly explained by weight allocaton seen above.


## Backtesting

```{r}
benchmark_returns <- Return.portfolio(R = returns,
                                      weights = equal_weights,
                                      rebalance_on = "years")

colnames(benchmark_returns) <-"benchmark"

base_port_spec <- port_spec

opt_base <- optimize.portfolio.rebalancing(R = returns,
                                           optimize_method = "ROI",
                                           portfolio = base_port_spec,
                                           rebalance_on = "quarters",
                                           training_period = 60,
                                           rolling_window = 60)

base_returns <- Return.portfolio(returns, extractWeights(opt_base))
colnames(base_returns) <- "base"

chart.Weights(opt_base, main = "Unconstrained Weights")


box_port_spec <- port_spec

box_port_spec <- add.constraint(portfolio = box_port_spec,
                                type = "box",
                                min = 0.05, max = 0.4,
                                indexnum = 2)

opt_box <- optimize.portfolio.rebalancing(R = returns,
                                         optimize_method = "ROI",
                                         portfolio = box_port_spec,
                                         rebalance_on = "quarters",
                                         training_period = 60,
                                         rolling_window = 60)

box_returns <- Return.portfolio(returns, extractWeights(opt_box))
colnames(box_returns) <- "box"

chart.Weights(opt_box, main = "Box Weights")


ret <- cbind(benchmark_returns, base_returns, box_returns)

table.AnnualizedReturns(ret)
```
Base Weights: unconstrained optimizer 
Box Weight: box constraints (min 5%, max 40% per asset)

- The **unconstrained weighted portfolio** shows extreme concentration patterns, often allocating nearly 100% to a single asset. This explains both the higher volatility and lower Sharpe ratio compared to the benchmark.

- The **box weighted portfolio** shows a more balanced distribution among all five assets throughout the period. This results in lower volatility than the base portfolio, improving the Sharpe ratio despite lower returns.

### Key Takeaways
- **Unconstrained optimization** can lead to extreme, concentrated bets, increasing risk and sometimes reducing risk-adjusted returns.
- **Box constraints** force diversification, lowering risk and often improving the Sharpe ratio, even if raw returns decrease.
- **Simple strategies** like equal-weighting can outperform more complex optimization in both return and risk-adjusted terms, especially when optimizer overfit or ignore diversification.
